---
title: "English Premier League Exploratory Analysis"
author: "Jai D. Bansal"
output: word_document
---

This report conducts exploratory analysis using English Premier League match data from the 2003/2004 season to the 2015/2016 season. The data can be obtained from: http://www.football-data.co.uk/englandm.php

```{r prep_import_and_clean, include = F, echo = F}
  
  # Load packages.
  library(data.table)
  library(qdapRegex)
  library(ggplot2)
  library(reshape2)

  # Import data.
  premier_data = data.table(read.csv('premier_data.csv', 
                                     header = T, 
                                     stringsAsFactors = F))

  # Change 'Date' to date format.
  # This was done in 'data_aggregation_and_cleaning.R' but must be done again since 'Date' gets imported as a 'character' vector.
  premier_data$Date = as.Date(premier_data$Date, 
                              '%Y-%m-%d')
  
  # Make all text fields lowercase and remove extra whitespace.
  premier_data$HomeTeam = rm_white(tolower(premier_data$HomeTeam))
  premier_data$AwayTeam = rm_white(tolower(premier_data$AwayTeam))
  premier_data$Referee = rm_white(tolower(premier_data$Referee))

  # Remove features from 'premier_data' and put them into 'feature_data'.
  # 'feature_data' is used in the 'Match Outcome Prediction' section.
  # I do this to remove clutter and allow the data to be viewed more easily.
  feature_data = premier_data[, .(FTR, home_team_wins, away_team_wins, home_avg_goals, away_avg_goals, home_avg_sot, away_avg_sot, home_avg_corner, away_avg_corner, home_avg_foul, away_avg_foul, home_avg_yellow, away_avg_yellow, home_avg_red, away_avg_red)]
  premier_data = premier_data[, .(Date, HomeTeam, AwayTeam, FTHG, FTAG, FTR, Referee, HS, AS, HST, AST, HC, AC, HF, AF, HY, AY, HR, AR, season_num)]


```

Below is a sample of the data and some summary statistics. I've removed columns that aren't used in the analysis (half time goals and results, total shots, and some betting odds).   
These might be used in future iterations.

### Data Sample:

```{r view_data, echo = F}

  # View data sample.
  head(premier_data)

```

The abbreviated columns stand for:
* FTHG: home team goals at end of match
* FTAG: away team goals at end of match
* FTR: match result ('h' denotes home team victory, 'a' denotes away team victory, 'd' denotes draw)
* HST: home team shots on target
* AST: away team shots on target
* HC: home team corner kicks
* AC: away team corner kicks
* HF: home team fouls
* AF: away team fouls
* HY: home team yellow cards
* AY: away team yellow cards
* HR: home team red cards
* AR: away team red cards

### Summary Statistics:

* First Match Date: `r min(premier_data$Date)`
* Last Match Date: `r max(premier_data$Date)`
* Number of Seasons: `r length(unique(premier_data$season_num))`
* Number of Matches: `r nrow(premier_data)`
* Number of Teams: `r length(unique(premier_data$HomeTeam))`
* Number of Referees: `r length(unique(premier_data$Referee))`
* % of Matches Won by Home Team: `r round(100 * nrow(premier_data[FTR == 'h']) / nrow(premier_data), 2)`
* % of Matches Won by Away Team: `r round(100 * nrow(premier_data[FTR == 'a']) / nrow(premier_data), 2)`
* % of Matches Ending in a Draw: `r round(100 * nrow(premier_data[FTR == 'd']) / nrow(premier_data), 2)`

### Relegation:

At the end of each season, the bottom 3 teams in the standings are relegated to a lower division.
The English Premier League (EPL) is the top English league, so teams cannot be 'promoted' out of the league.
I will look at how many seasons the teams in our dataset have spent in the EPL.

```{r relegation, echo = F}
  
    # Count seasons played by each team.
    premier_data[, 
                 seasons_played := length(unique(season_num)), 
                 by = 'HomeTeam']
  
    # Create unique table with 1 row per team.
    seasons = premier_data[, .(HomeTeam, seasons_played)]
    seasons = unique(seasons)
    
    # Create table with number of teams and how many seasons they played.
    seasons_result = data.table(table(seasons$seasons_played))
    setnames(seasons_result, 
             names(seasons_result), 
             c('Seasons', 'Teams'))
    
    # Show 'seasons_result'.
    print(seasons_result)
    
```

Out of `r length(unique(premier_data$HomeTeam))` teams in the data, 6 (`r round(100 * 6/38, 1)`%) only spent 1 year in the EPL.
8 teams (`r round(100 * 8/38, 1)`%) spent 13 years, the longest time possible in this data set.
The average number of seasons spent in the EPL is `r round(mean(seasons$seasons_played), 1)`.
Note that large numbers of teams played 13 seasons and 1 season, with relatively low numbers of teams for the number of seasons in between.

This data implies that there are many teams that break into the EPL from a lower division...and are promptly relegated the following year.
It also implies that there is a group of dominant teams that are rarely relegated.
Looking at the clubs that played all 13 seasons in the EPL (`r seasons[seasons_played == 13]$HomeTeam`) reveals many well-known (and wealthy) teams.

**Further Exploration**:

* Look for differences between teams that were promoted to the EPL and stayed there as opposed to teams that were promoted and then relegated the next year.

### Best of the Best and Worst of the Best:

In this section, I plot the win percentages over time of 4 teams (I only use 4 teams so the plot is readable). Win percentage is defined as (# of wins / # of matches). I only use teams that are in the data set for every season. This is an important bias. Teams that are in the data every season are consistently above a certain quality as they do not get relegated. Simply put, they are the best teams. I will plot the top 2 teams (best of the best) and the worst 2 teams (worst of the best).

```{r best_worst, fig.width = 6, fig.height = 4.5, echo = F}
   
    # Create 'winner' column in 'premier_data'. This helps with counting below.
    premier_data$winner = ifelse(premier_data$FTR == 'h', premier_data$HomeTeam, 
                                 ifelse(premier_data$FTR == 'a', premier_data$AwayTeam, 'draw'))
    
    # Create 'team_season_list' of all teams in all games specified by season. This helps with counting below.
    home_team_seasons = premier_data[, .(HomeTeam, season_num)]
    away_team_seasons = premier_data[, .(AwayTeam, season_num)]
    setnames(home_team_seasons, 
             names(home_team_seasons), 
             c('team', 'season_num'))
    setnames(away_team_seasons, 
             names(away_team_seasons), 
             c('team', 'season_num'))
    team_season_list = rbind(home_team_seasons, away_team_seasons)
    rm(home_team_seasons, away_team_seasons)

    # Restrict 'seasons' from above to only teams that played 13 seasons.
    seasons = seasons[seasons_played == 13]
    
    # Add 'total_wins','total_games', and a wins and games column for each season to 'seasons'. They're blank for now.
    seasons$total_win_percent = rep(NA, 
                                    nrow(seasons))
    seasons$s1_win_percent = rep(NA, 
                                 nrow(seasons))
    seasons$s2_win_percent = rep(NA, 
                                 nrow(seasons))
    seasons$s3_win_percent = rep(NA, 
                                 nrow(seasons))
    seasons$s4_win_percent = rep(NA, 
                                 nrow(seasons))
    seasons$s5_win_percent = rep(NA, 
                                 nrow(seasons))
    seasons$s6_win_percent = rep(NA, 
                                 nrow(seasons))
    seasons$s7_win_percent = rep(NA, 
                                 nrow(seasons))
    seasons$s8_win_percent = rep(NA, 
                                 nrow(seasons))
    seasons$s9_win_percent = rep(NA, 
                                 nrow(seasons))
    seasons$s10_win_percent = rep(NA, 
                                  nrow(seasons))
    seasons$s11_win_percent = rep(NA, 
                                  nrow(seasons))
    seasons$s12_win_percent = rep(NA, 
                                  nrow(seasons))
    seasons$s13_win_percent = rep(NA, 
                                  nrow(seasons))
    
    # Fill in all empty 'season' values.
    # I use a loop here (and few other times throughout the analysis).
    # Vectorized operations are much faster in R, but the structure of the data makes them challenging to implement.
    # The data set is small enough that running loops seems to be manageable.
    for (i in 1:nrow(seasons))
      {
        
        # Fill in win values.
        seasons$total_win_percent[i] = 100 * nrow(premier_data[winner == seasons$HomeTeam[i]]) / nrow(team_season_list[team == seasons$HomeTeam[i]])
        seasons$s1_win_percent[i] = 100 * nrow(premier_data[winner == seasons$HomeTeam[i] & season_num == 1]) / nrow(team_season_list[team == seasons$HomeTeam[i] & season_num == 1])
        seasons$s2_win_percent[i] = 100 * nrow(premier_data[winner == seasons$HomeTeam[i] & season_num == 2]) / nrow(team_season_list[team == seasons$HomeTeam[i] & season_num == 2])
        seasons$s3_win_percent[i] = 100 * nrow(premier_data[winner == seasons$HomeTeam[i] & season_num == 3]) / nrow(team_season_list[team == seasons$HomeTeam[i] & season_num == 3])
        seasons$s4_win_percent[i] = 100 * nrow(premier_data[winner == seasons$HomeTeam[i] & season_num == 4]) / nrow(team_season_list[team == seasons$HomeTeam[i] & season_num == 4])
        seasons$s5_win_percent[i] = 100 * nrow(premier_data[winner == seasons$HomeTeam[i] & season_num == 5]) / nrow(team_season_list[team == seasons$HomeTeam[i] & season_num == 5])
        seasons$s6_win_percent[i] = 100 * nrow(premier_data[winner == seasons$HomeTeam[i] & season_num == 6]) / nrow(team_season_list[team == seasons$HomeTeam[i] & season_num == 6])
        seasons$s7_win_percent[i] = 100 * nrow(premier_data[winner == seasons$HomeTeam[i] & season_num == 7]) / nrow(team_season_list[team == seasons$HomeTeam[i] & season_num == 7])
        seasons$s8_win_percent[i] = 100 * nrow(premier_data[winner == seasons$HomeTeam[i] & season_num == 8]) / nrow(team_season_list[team == seasons$HomeTeam[i] & season_num == 8])
        seasons$s9_win_percent[i] = 100 * nrow(premier_data[winner == seasons$HomeTeam[i] & season_num == 9]) / nrow(team_season_list[team == seasons$HomeTeam[i] & season_num == 9])
        seasons$s10_win_percent[i] = 100 * nrow(premier_data[winner == seasons$HomeTeam[i] & season_num == 10]) / nrow(team_season_list[team == seasons$HomeTeam[i] & season_num == 10])
        seasons$s11_win_percent[i] = 100 * nrow(premier_data[winner == seasons$HomeTeam[i] & season_num == 11]) / nrow(team_season_list[team == seasons$HomeTeam[i] & season_num == 11])
        seasons$s12_win_percent[i] = 100 * nrow(premier_data[winner == seasons$HomeTeam[i] & season_num == 12]) / nrow(team_season_list[team == seasons$HomeTeam[i] & season_num == 12])
        seasons$s13_win_percent[i] = 100 * nrow(premier_data[winner == seasons$HomeTeam[i] & season_num == 13]) / nrow(team_season_list[team == seasons$HomeTeam[i] & season_num == 13])
      
      }
    rm(i)
  
    # Order 'seasons' by 'total_win_percent'.
    seasons = seasons[order(total_win_percent, 
                            decreasing = T)]
    
    # Limit 'seasons' to top and bottom 2 teams.
    # Any more is hard to read on the plot.
    seasons = seasons[c(1:2, ((nrow(seasons) - 1):nrow(seasons))),]
    
    # Remove 'seasons_played' and 'total_win_percent' from 'seasons'. It's not needed.
    seasons$seasons_played = NULL
    seasons$total_win_percent = NULL
    
    # Rename 'seasons' columns for ease of plotting.
    setnames(seasons, 
             names(seasons),
             c('Team', '03-04', '04-05', '05-06', '06-07', '07-08', '08-09', '09-10', '10-11', '11-12', '12-13', '13-14', '14-15', '15-16'))
    
    # Rename 'seasons$Team' with proper team names.
    seasons$Team = c('Man. United', 'Chelsea', 'Everton', 'Aston Villa')
    
    # Reshape data for plotting.
    seasons_reshape = melt(seasons, 
                           id = 'Team')
    
    # Create plot.
    ggplot(data = seasons_reshape, 
           aes(x = variable, 
               y = value, 
               group = Team,
               color = Team)) +
      geom_line() +
      geom_point() +
      ylim(c(0, 100)) +
      ggtitle('Win Percentage of Selected Teams') +
      xlab('Season') +
      ylab('Win Percentage') +
      theme(axis.text.x = element_text(color = 'black', 
                                   size = 11, 
                                   angle = 45), 
      axis.text.y = element_text(color = 'black', 
                                   size = 13), 
      axis.title.x = element_text(size = 13),
      axis.title.y = element_text(size = 13,
                                  vjust = 1),
      axis.ticks = element_blank(),
      plot.title = element_text(face = 'bold', 
                                size = 13), 
      legend.title = element_text(size = 12),
      legend.text = element_text(size = 12))
     
```

All 4 of these teams are part of the dominant 8 that are never relegated in this data set. I find it surprising that Everton and Aston Villa (especially Aston Villa) are consistently below a win percentage of 50%. It seems that these clubs haven't done well recently, but haven't done poorly enough to be relegated. Finally, it's interesting that all 4 teams on the plot do worse, in some cases much worse, in the 2015-2016 season than the 2014-2015 season.

**Further Exploration**:

* Conduct this analysis with more teams.

### Clustering By Playing Style:

In this section, I will K-means to cluster all `r length(unique(premier_data$HomeTeam))` teams in the data by average fouls per game and average shots on net percentage per game. The latter quantity is computed by dividing the shots on net by total shots. I only cluster by two dimensions to allow easy visualization. Since different teams have spent different amounts of time in the EPL, it's important to use average quantities (and not totals).

```{r clustering, echo = F}
   
    # Clean up previous section artifacts.
  rm(seasons, team_season_list, seasons_reshape)
  
  # Create data table for clustering analysis with 1 row per team.
  cluster_data = data.table(team = unique(premier_data$HomeTeam), 
                            home_games = rep(NA, length(unique(premier_data$HomeTeam))), 
                            away_games = rep(NA, length(unique(premier_data$HomeTeam))),
                            home_fouls = rep(NA, length(unique(premier_data$HomeTeam))),
                            away_fouls = rep(NA, length(unique(premier_data$HomeTeam))),
                            home_shots_on_target = rep(NA, length(unique(premier_data$HomeTeam))),
                            away_shots_on_target = rep(NA, length(unique(premier_data$HomeTeam))),
                            home_shots = rep(NA, length(unique(premier_data$HomeTeam))),
                            away_shots = rep(NA, length(unique(premier_data$HomeTeam))))

  # Fill in 'cluster_data'.
  for (i in 1:nrow(cluster_data))
    {
    
      cluster_data$home_games[i] = nrow(premier_data[HomeTeam == cluster_data$team[i]])
      cluster_data$away_games[i] = nrow(premier_data[AwayTeam == cluster_data$team[i]])
      cluster_data$home_fouls[i] = sum(premier_data[HomeTeam == cluster_data$team[i]]$HF) 
      cluster_data$away_fouls[i] = sum(premier_data[AwayTeam == cluster_data$team[i]]$AF)
      cluster_data$home_shots_on_target[i] = sum(premier_data[HomeTeam == cluster_data$team[i]]$HST)
      cluster_data$away_shots_on_target[i] = sum(premier_data[AwayTeam == cluster_data$team[i]]$AST)
      cluster_data$home_shots[i] = sum(premier_data[HomeTeam == cluster_data$team[i]]$HS)
      cluster_data$away_shots[i] = sum(premier_data[AwayTeam == cluster_data$team[i]]$AS)
  
  }
  rm(i)
  
  # Compute total games, fouls, shots on target, and shots (home quantity + away quantity).
  cluster_data$games = cluster_data$home_games + cluster_data$away_games
  cluster_data$fouls = cluster_data$home_fouls + cluster_data$away_fouls
  cluster_data$shots_on_target = cluster_data$home_shots_on_target + cluster_data$away_shots_on_target
  cluster_data$shots = cluster_data$home_shots + cluster_data$away_shots
  
  # Compute average fouls per game and average shot percentage per game for each team.
  cluster_data$avg_fouls = cluster_data$fouls / cluster_data$games
  cluster_data$avg_shots_on_target_percent = 100 * cluster_data$shots_on_target / cluster_data$shots
  
  # Scale 'avg_fouls' and 'avg_shots_on_target_percent'. This should be done prior to clustering.
  # I do this separately so I can later access the mean and standard deviation of scaling.
  avg_fouls_scaled = scale(cluster_data$avg_fouls)
  avg_shots_on_target_percent_scaled = scale(cluster_data$avg_shots_on_target_percent)
  
  # Add 'avg_fouls_scaled' and 'avg_shots_on_target_percent_scaled' to 'cluster_data'.
  cluster_data$avg_fouls_scaled = avg_fouls_scaled
  cluster_data$avg_shots_on_target_percent_scaled = avg_shots_on_target_percent_scaled
  
  # Conduct K-means clustering with 4 clusters.
  cluster_kmeans = kmeans(cluster_data[, .(avg_fouls_scaled, avg_shots_on_target_percent_scaled)], 
                          centers = 4)
  
  # Convert cluster centers back into unscaled values.
  cluster_centers = data.table(cluster_kmeans$centers)
  cluster_centers$avg_fouls_scaled = (cluster_centers$avg_fouls_scaled * attr(avg_fouls_scaled, 'scaled:scale')) + 
                                      attr(avg_fouls_scaled, 'scaled:center')
  cluster_centers$avg_shots_on_target_percent_scaled = (cluster_centers$avg_shots_on_target_percent_scaled * attr(avg_shots_on_target_percent_scaled, 'scaled:scale')) + attr(avg_shots_on_target_percent_scaled, 'scaled:center')
  
  # Add clusters to 'cluster_data'.
  cluster_data$clusters = cluster_kmeans$cluster
  
  # View cluster means.
  cluster_centers$avg_fouls_scaled = round(cluster_centers$avg_fouls_scaled, 1)
  cluster_centers$avg_shots_on_target_percent_scaled = round(cluster_centers$avg_shots_on_target_percent_scaled, 1)
  setnames(cluster_centers, names(cluster_centers), c('Average Fouls', 'Avg. Shots on Net %'))
  cluster_centers
  
  # Plot clusters.
  ggplot(data = cluster_data[, .(avg_fouls, avg_shots_on_target_percent, clusters)], 
         aes(x = avg_fouls, 
             y = avg_shots_on_target_percent, 
             group = as.character(clusters),
             color = as.character(clusters))) +
    geom_point(size = 2, 
               aes(shape = as.character(clusters))) +
    scale_color_manual(values = c('red', 'blue', 'yellow', 'black')) +
    theme(axis.text.x = element_text(color = 'black', 
                                     size = 13), 
          axis.text.y = element_text(color = 'black', 
                                     size = 13), 
          axis.title.x = element_text(face = 'bold', 
                                    size = 13),
          axis.title.y = element_text(face = 'bold',
                                      size = 13, 
                                      vjust = 1),
          axis.ticks = element_blank(),
          plot.title = element_text(face = 'bold', 
                                    size = 13)) +
    guides(colour = F, 
           shape = F) +
    ggtitle('Clustered EPL Teams') +
    xlab('Average Fouls Per Game') +
    ylab('Avg. Shots on Net % Per Game')
 
```

Oddly enough, RMarkdown output has different colors/shapes in different positions everytime This needs to be filled in properly in the final version. The  <> cluster represents teams with many fouls and relatively high shot accuracy. The <> cluster commits less fouls than the blue cluster and has a slightly lower accuracy. The <> cluster is characterized by a lower shot accuracy. The <> cluster admittedly looks like a miscellaneous cluster, but is generally characterized by lower shot accuracy and fouls.

**Further Exploration**:

* Add more dimensions to the clustering.

### Match Outcome Prediction:

In this section, I predict the outcome of matches (home win, away win, draw).   
I use a variety of features (wins, average goals per match, average on-target shots per match, average corner kicks per match, average fouls per match, average yellow cards per match, average red cards per match) from the **same season** only.   
Each feature is computed for the home and away teams of the match.
For example, suppose team A is playing team B in season 2. I would compute all features for both teams using only previous matches in season 2.   
This method means that I cannot obtain predictions for a team's 1st match of the season.  
For a team's 2nd match of the season, I will obtain predictions using only data from that team's 1st match.
For a team's 10th match, I obtain predictions using data from the 1st 9 matches.   
So, each team's features are being updated every match.

```{r match_outcome_prediction, echo = F}
  
  # Remove unneeded objects:
  rm(avg_fouls_scaled, avg_shots_on_target_percent_scaled, cluster_centers, cluster_data, seasons_result, cluster_kmeans)

```
   